Temporal ordering and registration of images in studies of developmental dynamics

### 2014-05-29 image_analysis_paper/paper_figures2/fig5.tex
edited paper

%\usepackage[margin=0in, paperwidth=8.7cm, paperheight=9.5cm]{geometry}
\usepackage[margin=0in, paperwidth=8.7cm, paperheight=16.5cm]{geometry}

@@

\centering
%%\raisebox{5.4cm}{\textsf{A}}
%\begin{tikzpicture}
%     \node at (0,0)
%       {\includegraphics[width=8cm]{mut_wt_vdm_embedding_background}};
%     \node at (0,0)
%       {\includegraphics[width=8cm]{mut_wt_vdm_embedding}};
%      \node at (-4cm, 2.4cm) {\textsf{A}};
%\end{tikzpicture}\\
\raisebox{5.2cm}{\textsf{A}}
\includegraphics[width=6cm, angle=90]{data3_embed_bw}\\
\raisebox{5.2cm}{\textsf{B}}
\includegraphics[width=8cm]{data3_embed_color}\\
\begin{tikzpicture}
	\node at (-4.25cm, 0cm) {\textsf{C}};
	\node at (0,0) {\includegraphics[width=8cm]{wt_trajectory}};
	%\draw[help lines,step=.5] (-4,-1) grid (4,1);
	\draw[->, color=white, thick] (-1.1cm, -0.1cm) -- (-1.3cm, -0.3cm);
	\draw[->, color=white, thick] (-0.7cm, -0.1cm) -- (-0.5cm, -0.2cm);

>>>>

%\usepackage[margin=0in, paperwidth=8.7cm, paperheight=9.5cm]{geometry}
\usepackage[margin=0in, paperwidth=8.7cm, paperheight=17cm]{geometry}

@@

\centering
\begin{tikzpicture}
	\node at (0cm, 0cm) {\includegraphics[width=6cm, angle=90]{data3_embed_bw}};
	\node at (-4cm, 2.5cm) {\textsf{A}};
\end{tikzpicture}\\
\begin{tikzpicture}
	\node at (0cm, 0cm) {\includegraphics[width=8cm]{data3_embed_color}};
	\node at (-4cm, 2.5cm) {\textsf{B}};
	%\draw[help lines,step=.5] (-4,-4) grid (4,4);
	\draw [draw=black!20!green, ->, ultra thick] (-2, -2) .. controls (-1.1, 0.2) and  (1, 0)  .. (2, -2);
	\draw [blue, ->, ultra thick] (-2.65, -1.8)  .. controls (-2.3, 0.9) and (-1.7, 1.5) .. (-1.4, 2.1); 
\end{tikzpicture}\\
\begin{tikzpicture}
	\node at (-4.25cm, 0cm) {\textsf{C}};
	\node at (0,0) {\includegraphics[width=8cm]{wt_trajectory}};
	%\draw[help lines,step=.5] (-4,-1) grid (4,1);
	\draw[->, color=white, thick] (-1.1cm, -0.1cm) -- (-1.3cm, -0.3cm);
	\draw[->, color=white, thick] (-0.7cm, -0.1cm) -- (-0.5cm, -0.2cm);

### 2014-05-29 image_analysis_paper/pnas_draft2.tex
edited paper

We show how recently developed dimensionality reduction techniques can successfully address these challenges and reveal developmental trajectories from collections of unregistered and unordered imaging data.
%
Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}, as have dimensionality reduction techniques \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear} and their applications \cite{anavy2014blind, trapnell2014dynamics,gupta2008extracting, qiu2011discovering, kafri2013dynamics}. 
%
Here, we use a manifold learning algorithm \cite{singer2012vector} that has been recently developed for high-dimensional imaging data. 
%
This technique has been used for data mining in
applications ranging from analysis of cryo-electron microscopy images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of imaging studies. 
%
Here it is adapted for simultaneous registration and temporal ordering of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
%
We analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
%
%The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate this approach.
%
%In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
%
%Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
%
%The third set consists of data collected from both wild type and mutant embryos, where joint processing of the data will help to visualize the different sources of variability along distinct developmental paths.

@@

For this data set, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
%
During the third hour of development, the embryo cellularizes: in this process, nuclei located under a common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
%
The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 1--3 minutes \cite{figard2013plasma} (see \fig~S1).
%
The Spearman rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.91, indicating that vector diffusion maps can accurately recover the temporal order.
%
Thus, we have validated a dimensionality reduction approach to temporal ordering and registration of data on a data set where image orientation and ordering can be assessed independently.

@@

%As the images in Figure~\ref{fig:data1} are relatively simple, other techniques could have been used to temporally order them.
%
Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~S6).
%
However, developmental dynamics are often significantly more complex, and nonlinear dimensionality reduction techniques are required to extract meaningful structure in the data.
%

@@

Visually, the dynamics in this data set are much more complicated, and only 22\% of the variability in the images is captured by the first principal component.
%
Simple time-ordering based on this linear projection is therefore not possible (see \fig~S7).
%
%Even when projecting on several leading principal components (see SI), a nonlinear approach, such as vector diffusion maps, is required to temporally order the data.
%
\fig~\ref{subfig:ordered_data2} shows the images in \fig~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
%
Visual inspection confirms this ordering is consistent with the known developmental dynamics outlined above.
%
However, assimilating the dynamics even from this concatenation of images is difficult for two reasons, even after registration and ordering.
%
First, the sheer number of images makes visual processing of the entire data set nontrivial, and
second, the ordered data is not entirely smooth due to inter-embryo variability.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path.  
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:averaged_data2}, serve as a summary of a stereotypic developmental dynamics reconstructed from cross-sectional data.
%
Thus, vector diffusion maps can accomplish the image registration and ordering tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).

@@

When time is the main source of variability, we have shown vector diffusion maps can recover developmental progression from cross-sectional data.
%
However, vector diffusion maps can also provide an organization of data which contains multiple sources of variability.
%
To illustrate this, we collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see \fig~S2).
%
Half of the embryos are wild type, and half of the embryos are mutants where dpERK is expressed in one broad peak at the ventral side.
%
Again, each image is an optical cross-section of an embryo fixed at a different developmental time, and the embryos have been stained for nuclei (gray), Dl (green), and dpERK (red).
%
We expect two sources of variability in this data set: differences due to developmental time, and differences between mutant and wild type dynamics.

\begin{figure}[h!]
\includegraphics{fig5}
\caption{Mutant and wild type {\xfigtextfontit Drosophila}.  {\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~S2), projected onto the first two vector diffusion maps coordinates. {\xfigtextfontit (B)} The data in {\xfigtextfontit A}, now colored by genotype. The two branches separate mutant and wild type samples. {\xfigtextfontit (C)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by circles). The two dpERK peaks which arise during development are indicated by arrows. {\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by circles). The single dpERK peak is indicated by an arrow. }
\customlabel{fig:data3}{5}
\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
\customlabel{subfig:projection_colored_data3}{\ref{fig:data3}{\it B}}
\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it C}}
\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it D}}
\end{figure}

We computed the vector diffusion maps embedding for our data, and the eigenvalue spectrum (see \fig~S5) indicates that two coordinates may be required for a thorough description of the data. 
%
\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates, where each point in this projection corresponds to one image in our data set.
%%%
%%%YGK we need something about dmap eignvalues and how many are not harmonics
%
Two distinct branches appear to emerge in this two-dimensional embedding space.

@@

@@

The presented approach was demonstrated on three different experimental data sets, collected specifically for this study. 
%
In the first example, correct orientation and temporal order of images was known {\it a priori}, providing a test case for the algorithm in which both of these features were accurately recovered. 
%
For this particular data set, a simpler approach would have sufficed. 
%
Indeed, because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern and relatively small tissue deformation), temporal order could have been inferred by projection of registered images on the first principal component of the data set. 
%
This approach was clearly not sufficient for the second data set (see \fig~S3), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.  
%
However, vector diffusion maps could readily extract a nontrivial spatiotemporal trajectory. 
%
Finally, the algorithm was applied to a heterogeneous data set, composed of images from embryos of different genotypes. 
%
In this case, low-dimensional projection of registered data revealed distinct paths that correlated with the developmental trajectories of the mutant and wild type populations.

>>>>

Image registration has been studied in a variety of contexts \cite{zitova2003image, rowley1998rotation, hajnal2010medical, greenspan1994rotation, zhao2003face}.
%
We use a manifold learning algorithm \cite{singer2012vector} which simultaneously addresses the problems of registration and temporal ordering. 
%
This algorithm is one of many nonlinear dimensionality reduction techniques that have been developed over the past decade \cite{Belkin2003, coifman2005geometric, coifman2006geometric, tenenbaum2000global, roweis2000nonlinear}. 
%
This technique has been used for data mining in
applications ranging from analysis of cryo-electron microscopy images of individual molecules to face recognition, and is sufficiently general to be applicable to a wide variety of imaging studies. 
%
Here it is adapted for simultaneous registration and temporal ordering of images from cross-sectional studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories.
%
We analyze three data sets from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of developmental dynamics \cite{jaeger2012drosophila}.
%
%The first data set, shown in Figure~\ref{subfig:raw_data1}, is relatively simple and will allow us to both illustrate and validate this approach.
%
%In the second data set, shown in Figure~\ref{subfig:raw_data2}, the dynamics are significantly more complex.
%
%Here, we will show that we can recover a trajectory that is qualitatively consistent with current knowledge.
%
%The third set consists of data collected from both wild type and mutant embryos, where joint processing of the data will help to visualize the different sources of variability along distinct developmental paths.

@@

For this data set, the quality of the temporal ordering provided by vector diffusion maps can be assessed by comparing with the ordering obtained through an independent approach.
%
During the third hour of development, the embryo cellularizes: in this process, nuclei located under a common plasma membrane are enclosed by lateral membranes which grow inward, separating the nuclei into individual cells.
%
The highly reproducible kinetics of lateral membrane growth can be used to estimate the age of each embryo to within 1--3 minutes \cite{figard2013plasma} (see \fig~S2).
%
The Spearman rank correlation coefficient between the orderings obtained based on the progress of cellularization and vector diffusion maps is 0.91, indicating that vector diffusion maps can accurately recover the temporal order.
%
Thus, we have validated a dimensionality reduction approach to temporal ordering and registration of data on a data set where image orientation and ordering can be assessed independently.

@@

%As the images in Figure~\ref{fig:data1} are relatively simple, other techniques could have been used to temporally order them.
%
Because the dynamics of the data set shown in \fig~\ref{fig:data1} are relatively simple, effectively consisting of a single mode (the two dpERK peaks) growing in time, established techniques such as principal component analysis (PCA) \cite{shlens2005tutorial} would uncover most of the meaningful structure in the data, and projection onto the first principal component would be sufficient to order the data in time (see \fig~S7).
%
However, developmental dynamics are often significantly more complex, and nonlinear dimensionality reduction techniques are required to extract meaningful structure in the data.
%

@@

Visually, the dynamics in this data set are much more complicated, and only 22\% of the variability in the images is captured by the first principal component.
%
Simple time-ordering based on this linear projection is therefore not possible (see \fig~S8).
%
%Even when projecting on several leading principal components (see SI), a nonlinear approach, such as vector diffusion maps, is required to temporally order the data.
%
\fig~\ref{subfig:ordered_data2} shows the images in \fig~\ref{subfig:raw_data2}, now registered and ordered using vector diffusion maps \cite{singer2012vector}.
%
Visual inspection confirms this ordering is consistent with the known developmental dynamics outlined above.
%
However, assimilating the dynamics even from this concatenation of images is difficult for two reasons, even after registration and ordering.
%
First, the sheer number of images makes visual processing of the entire data set nontrivial, and
second, the ordered data is not entirely smooth due to inter-embryo variability.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we applied a Gaussian filter to the trajectory of registered and ordered images to obtain an averaged developmental path.  
%
Sequential snapshots of this averaged trajectory, shown in \fig~\ref{subfig:averaged_data2}, serve as a summary of a stereotypic developmental dynamics reconstructed from cross-sectional data.
%
Thus, vector diffusion maps can accomplish the image registration and ordering tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental dynamics (e.g., correlation of age with body size).

@@

When time is the main source of variability, we have shown vector diffusion maps can recover developmental progression from cross-sectional data.
%
However, vector diffusion maps can also provide an organization of data which contains multiple sources of variability.
%
To illustrate this, we collected a data set consisting of $42$ embryos undergoing cellularization during the third hour of development (see \fig~S3).
%
Half of the embryos are wild type, and half of the embryos are mutants where dpERK is expressed in one broad peak at the ventral side.
%
Again, each image is an optical cross-section of an embryo fixed at a different developmental time, and the embryos have been stained for nuclei (gray), Dl (green), and dpERK (red).
%
We expect two sources of variability in this data set: differences due to developmental time, and differences between mutant and wild type dynamics.

We computed the vector diffusion maps embedding for our data, and the eigenvalue spectrum (see \fig~S6) indicates that two coordinates may be required for a thorough description of the data. 
%
\fig~\ref{subfig:projection_data3} shows the data projected onto the first two vector diffusion maps coordinates, where each point in this projection corresponds to one image in our data set.
%%%
%%%YGK we need something about dmap eignvalues and how many are not harmonics
%
Two distinct branches appear to emerge in this two-dimensional embedding space.

@@

\begin{figure}[h!]
\includegraphics{fig5}
\caption{Mutant and wild type {\xfigtextfontit Drosophila}.  {\xfigtextfontit(A)} Data collected from wild type and mutant embryos (original data set is shown in \fig~S3), projected onto the first two vector diffusion maps coordinates. {\xfigtextfontit (B)} The data in {\xfigtextfontit A}, now colored by genotype. The two branches separate mutant and wild type samples, and the direction of the developmental trajectories are indicated by arrows. {\xfigtextfontit (C)} dpERK signal at select wild type points in {\xfigtextfontit B} (denoted by circles). The two dpERK peaks which arise during development are indicated by arrows. {\xfigtextfontit (C)} dpERK signal at select mutant points in {\xfigtextfontit B} (denoted by circles). The single dpERK peak is indicated by an arrow. }
\customlabel{fig:data3}{5}
\customlabel{subfig:projection_data3}{\ref{fig:data3}{\it A}}
\customlabel{subfig:projection_colored_data3}{\ref{fig:data3}{\it B}}
\customlabel{subfig:average_wt_data3}{\ref{fig:data3}{\it C}}
\customlabel{subfig:average_mut_data3}{\ref{fig:data3}{\it D}}
\end{figure}

@@

The presented approach was demonstrated on three different experimental data sets, collected specifically for this study. 
%
In the first example, correct orientation and temporal order of images was known {\it a priori}, providing a test case for the algorithm in which both of these features were accurately recovered. 
%
For this particular data set, a simpler approach would have sufficed. 
%
Indeed, because of the relative simplicity of developmental dynamics (monotonic increase in the amplitude of a localized signaling pattern and relatively small tissue deformation), temporal order could have been inferred by projection of registered images on the first principal component of the data set. 
%
This approach was clearly not sufficient for the second data set (see \fig~S4), collected from embryos during the time window where both the tissue morphology and signaling patterns are rapidly changing.  
%
However, vector diffusion maps could readily extract a nontrivial spatiotemporal trajectory. 
%
Finally, the algorithm was applied to a heterogeneous data set, composed of images from embryos of different genotypes. 
%
In this case, low-dimensional projection of registered data revealed distinct paths that correlated with the developmental trajectories of the mutant and wild type populations.

### 2014-05-29 image_analysis_paper/si_information.tex
edited paper

To find this coordinate $y$, we want solve the following optimization problem \cite{Belkin2003}
\begin{equation} \label{eq:dmaps_opt_problem}
\argmin_{y} \sum_{ij} W_{ij} (y(x_i) - y(x_j))^2.
\end{equation}
%
We first compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
\begin{equation} \label{eq:dmaps_A}
A = D^{-1} W.
\end{equation}
%
We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
%
Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.
%
Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector; this is a trivial solution to \eqref{eq:dmaps_opt_problem}.
%
%In general, the next few eigenvectors $\phi_2, \dots, \phi_m$ give ``meaningful'' embedding coordinates for the data, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point.
%
The next eigenvector, $\phi_2$, is the (non-trivial) solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)$, the $j^{th}$ entry of $\phi_2$, gives the ``new" coordinate for data point $x_j$ (i.e., $\phi_2(j) = y(x_j)$).
%
In our application, we have assumed that this single direction of variability, parameterized by $\phi_2$, is one-to-one with time.
%
Ordering the data by $\phi_2(j)$ will then, effectively, order them in time.
%
The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space, where lower eigenvectors give subsequent embedding coordinates for the data.
%
The magnitude of the eigenvalues, $|\lambda_i|$, gives a measure of ``importance'' of the corresponding embedding coordinate $\phi_i$.
%
Often, one looks for a spectral gap in the eigenvalues to determine how many dimensions are necessary to adequately describe the data.

@@

In general, the embedding coordinates are given by
\begin{equation} \label{eq:vdm_coord}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle,
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$, 
%
If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate.
%
As the first diffusion maps coordinate is constant, the first block eigenvector contains only the optimal rotations.
%
The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate. 
%
We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
%
The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is the solution to
$$ \max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. $$
%
As in diffusion maps, the magnitude of the eigenvalues is related to the importance of the coordinates and gives an indication of the inherent dimensionality of the data. 
%
Again, if the rotations and dynamic variation are separable, then we expect the each eigenvalue to appear with multiplicity $d$, as each block eigenvector will yield $d$ eigenvalues of approximately the same value. 
%
We therefore look for a spectral gap in the ``blocks'' of eigenvalues to determine how many components are necessary to describe the data.

@@

@@

Once we have registered and ordered the images, we smooth the resulting trajectory to obtain a movie of the average developmental dynamics. 
%
Let $x_1, \dots, x_m$ denote the set of {\em registered and ordered} images (so $x_i$ is the $i^{th}$ image in the ordered set).
%
We define the frame at time $\tau$, denoted $\overline{x}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
\begin{equation}
\overline{x}_{\tau} =  \frac{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) x_i}{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) }
\end{equation}
where $\sigma$ is the scale of the Gaussian filter. 
%
For the images in \fig~4, we take $\sigma^2 = 20$.

@@

@@

\begin{figure*}
\raisebox{4.5cm}{\sffamily{A}}
\includegraphics[width=5cm]{data1_evals}
\raisebox{4.5cm}{\sffamily{B}}
\includegraphics[width=5cm]{data2_evals}
\raisebox{4.5cm}{\sffamily{C}}
\includegraphics[width=5cm]{data3_evals}

\caption{The eigenvalue spectra from vector diffusion maps analysis for the three data sets. 
%
As expected, the eigenvalues appear in groups of 3, as $d=3$ when using vector diffusion maps to register with respect to translations and rotations. 
%
(A) Spectrum from images during cellularization (\fig~3). As expected, there is a spectral gap after the first (non-trivial) group of eigenvalues: the first three eigenvalues correspond to the rotational and translational alignment, the second three eigenvalues correspond to the first VDM component which uncovers the developmental dynamics, and the then there is an appreciable gap before the third group of eigenvalues.
%
(B) Spectrum from images during gastrulation (\fig~4). Again, 
there is a spectral gap after the first (non-trivial) group of eigenvalues.
%
(C) Spectrum from images of mutant and wild type embryos during cellularization (\fig~5).
We see a group of three eigenvalues corresponding to the translational and rotational alignments, one group that is clearly significant, and then a second group that, while less important, is still above the level of noise of the subsequent eigenvalues, indicating that two components are necessary to describe the data. }
\label{fig:eval_spectra}
\end{figure*}

>>>>

To find this coordinate $y$, we want solve the following optimization problem \cite{Belkin2003}
\begin{equation} \label{eq:dmaps_opt_problem}
\argmin_{y} \sum_{ij} W_{ij} (y(x_i) - y(x_j))^2.
\end{equation}
%
We first compute the diagonal matrix $D$, where $D_{ii} = \sum_{j=1}^{m} W_{ij}$, and the matrix $A$, where
\begin{equation} \label{eq:dmaps_A}
A = D^{-1} W.
\end{equation}
%
We calculate the eigenvectors $\phi_1, \phi_2, \dots, \phi_m$, ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
%
Because the matrix $A$ is similar to the symmetric matrix $D^{-1/2} W D^{-1/2}$, $A$ is guaranteed to have real eigenvalues and real, orthogonal eigenvectors.
%
Because the matrix $A$ is row-stochastic, $\lambda_1=1$ and $\phi_1$ is a constant vector; this is a trivial solution to \eqref{eq:dmaps_opt_problem}.
%
%In general, the next few eigenvectors $\phi_2, \dots, \phi_m$ give ``meaningful'' embedding coordinates for the data, such that $\phi_j(i)$ gives the $j^{th}$ embedding coordinate of the $i^{th}$ data point.
%
The next eigenvector, $\phi_2$, is the (non-trivial) solution to \eqref{eq:dmaps_opt_problem}, so that $\phi_2(j)$, the $j^{th}$ entry of $\phi_2$, gives the ``new" coordinate for data point $x_j$ (i.e., $\phi_2(j) = y(x_j)$).
%
In our application, we have assumed that this single direction of variability, parameterized by $\phi_2$, is one-to-one with time.
%
Ordering the data by $\phi_2(j)$ will then, effectively, order them in time.
%
The procedure generalizes when the data lie on higher-dimensional manifolds (not just curves) in data space, where lower eigenvectors give subsequent embedding coordinates for the data.
%

@@

In general, the embedding coordinates are given by
\begin{equation} \label{eq:vdm_coord}
\psi_{k,l} (i) = \langle \phi_k(i), \phi_l(i) \rangle,
\end{equation}
where $\phi_k(i) \in \mathbb{R}^d$ denotes the $i^{th}$ block of $\phi_k$, 
%
If we assume that the rotations and the dynamics are uncoupled and therefore separable, then the eigenvectors of $S$ have the following structure: each block eigenvector contains estimates of the optimal rotations (up to a constant rotation) multiplied by the corresponding embedding coordinate.
%
As the first diffusion maps coordinate is constant, the first block eigenvector contains only the optimal rotations.
%
The second block eigenvector (eigenvectors $d+1$ through $2d$) contains the optimal rotations, each multiplied by their second diffusion maps coordinate. 
%
We can therefore recover this diffusion maps coordinate by taking inner products of the columns of the second block eigenvector with columns of the first block eigenvector.
%
The $j^{th}$ embedding coordinate will be given by $\psi_{k,1}$, where $jd +1 < k \le (j+1)d$,
and we select $k$ such that the coordinate $\psi_{k, 1}$ has the largest variability, i.e., the $j^{th}$ coordinate is $\psi_{k,1}$, where $k$ is the solution to
$$ \max_{jd +1 < k \le (j+1)d} \sum_i \psi_{k,1} (i)^2. $$

@@

\subsection{Determining dimensionality from eigenvalue spectrum}

We can use the distribution of the eigenvalues from (vector) diffusion maps to help determine the dimensionality of the data.
%
For diffusion maps, the largest eigenvalue will always be 1 and correspond to the trivial (constant) eigenvector. 
%
The eigenvalues corresponding to noise in the data form a semicircle distribution centered at 0. 
%
The eigenvalues corresponding to descriptive/meaningful embedding coordinates will be positive and removed from the semicircle. 
%
For vector diffusion maps, the sam structure holds, except each coordinate gives rise to $d$ eigenvalues with (approximately) the same value (as each block eigenvector corresponds to $d$ eigenvalues).

@@

Once we have registered and ordered the images, we smooth the resulting trajectory to obtain a picture of the stereotypic developmental dynamics. 
%
Let $x_1, \dots, x_m$ denote the set of {\em registered and ordered} images (so $x_i$ is the $i^{th}$ image in the ordered set).
%
We define the average image at time $\tau$, denoted $\overline{x}_{\tau}$ (where $1 \le \tau \le m$), as a (Gaussian) weighted average of the images,
\begin{equation}
\overline{x}_{\tau} =  \frac{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) x_i}{\sum_i \exp \left( - \frac{|i - \tau|^2}{\sigma^2} \right) }
\end{equation}
where $\sigma$ is the scale of the Gaussian filter. 
%
For the images in \fig~4, we take $\sigma^2 = 20$.

@@

\begin{figure}
\includegraphics[width=8.4cm]{ap_dv}
\caption{(Left) A lateral view of a {\it Drosophila} embryo stained with DAPI (gray), dpERK (red), and Dorsal (green). Embryo is presented so that the anterior (A) side is to the left and the posterior (P) side is to the right. The arrow indicates the position where the cross-section of an embryo is imaged. (Right) A dorsoventral view of a {\it Drosophila} embryo. The dorsal (D) side is up and the ventral (V) side is down. Images were collected at the focal plane ~18\% from either the anterior or posterior pole of an embryo (arrow in the left image). }
\end{figure}

@@

\begin{figure*}
%\raisebox{4.5cm}{\sffamily{A}}
%\includegraphics[width=5cm]{data1_evals}
\begin{tikzpicture}
	\node at (0,0) {\includegraphics[width=5cm]{data1_evals_dist}};
	\node at (-2.5cm, 2cm) {\sffamily{A}};
	\draw[<->, ultra thick, cyan] (-1.4cm, -1.9cm) -- (1.7cm, -1.9cm);
	\draw[->, ultra thick, red] (1.97cm, -0.5cm) -- (1.97cm, -1cm);
\end{tikzpicture}
%\raisebox{4.5cm}{\sffamily{B}}
%\includegraphics[width=5cm]{data2_evals}
\begin{tikzpicture}
	\node at (0,0) {\includegraphics[width=5cm]{data2_evals_dist}};
	\node at (-2.5cm, 2cm) {\sffamily{B}};
	\draw[<->, ultra thick, cyan] (-1.2cm, -1.9cm) -- (1.5cm, -1.9cm);
	\draw[->, ultra thick, red] (1.97, -0.85cm) -- (1.97cm, -1.34cm);
\end{tikzpicture}
%\raisebox{4.5cm}{\sffamily{C}}
%\includegraphics[width=5cm]{data3_evals}
\begin{tikzpicture}
	\node at (0,0) {\includegraphics[width=5cm]{data3_evals_dist}};
	\node at (-2.5cm, 2cm) {\sffamily{C}};
	\draw[<->, ultra thick, cyan] (-0.25cm, -1.9cm) -- (0.35cm, -1.9cm);
	\draw[->, ultra thick, red] (1.97, -1.1cm) -- (1.97cm, -1.6cm);
	\draw[->, ultra thick, red] (1.52, -1.1cm) -- (1.52cm, -1.6cm);
	\draw[->, ultra thick, red] (0.64, -1.1cm) -- (0.64cm, -1.6cm);
\end{tikzpicture}
\caption{The histograms of the eigenvalues from vector diffusion maps analysis for the three data sets. 
We expect to see an approximately symmetric semicircle distribution centered at 0, which corresponds to noise, and positive eigenvalues separated from the semicircle which correspond to meaningful coordinates. 
Each coordinate yields $d=3$ eigenvalues of (approximately) the same magnitude, and the top $d=3$ eigenvalues ($\lambda=1$) correspond to a trivial embedding coordinate.
(A) Histogram of eigenvalues from images during cellularization (\fig~3). As expected, we see an approximately symmetric distribution that ranges from $-0.8$ to $0.8$ (indicated by the blue arrow), and $6$ eigenvalues that are outside of this distribution (indicated by the red arrow). $3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first VDM component which uncovers the developmental dynamics, and the then there is an appreciable gap before the third group of eigenvalues.
(B) Histogram of eigenvalues from images during gastrulation (\fig~4). We see an approximately symmetric distribution that ranges from $-0.7$ to $0.7$ (indicated by the blue arrow), and $6$ eigenvalues that are outside of this distribution (indicated by the red arrow). $3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $3$ eigenvalues correspond to the first VDM component which uncovers the developmental dynamics.
(C) Histogram of eigenvalues from images of mutant and wild type embryos during cellularization (\fig~5).
We see an approximately symmetric distribution that ranges from $-0.2$ to $0.2$ (indicated by the blue arrow), and $9$ eigenvalues that are outside of this distribution (indicated by the red arrows). $3$ of these eigenvalues correspond to the eigenvector which contains the rotational and translational alignment, and the other $2$ eigenvalues correspond to the first two VDM components which uncovers the developmental dynamics of the two genotypes.}
%there is a spectral gap after the first (non-trivial) group of eigenvalues.
%\caption{The eigenvalue spectra from vector diffusion maps analysis for the three data sets. 
%%
%As expected, the eigenvalues appear in groups of 3, as $d=3$ when using vector diffusion maps to register with respect to translations and rotations. 
%%
%(A) Spectrum from images during cellularization (\fig~3). As expected, there is a spectral gap after the first (non-trivial) group of eigenvalues: the first three eigenvalues correspond to the rotational and translational alignment, the second three eigenvalues correspond to the first VDM component which uncovers the developmental dynamics, and the then there is an appreciable gap before the third group of eigenvalues.
%%
%(B) Spectrum from images during gastrulation (\fig~4). Again, 
%there is a spectral gap after the first (non-trivial) group of eigenvalues.
%%
%(C) Spectrum from images of mutant and wild type embryos during cellularization (\fig~5).
%We see a group of three eigenvalues corresponding to the translational and rotational alignments, one group that is clearly significant, and then a second group that, while less important, is still above the level of noise of the subsequent eigenvalues, indicating that two components are necessary to describe the data. }
\label{fig:eval_spectra}
\end{figure*}
